---
title: "bio_csus_uw"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
#install.packages("")
```

```{r, include = FALSE}
library(purrr)
library(scales)
library(plyr)
library(readtext)
library(stringr)
library(quanteda)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(openxlsx)
library(alluvial)
library(ggalluvial)
library(data.table)
library(ggrepel)
library(ggfittext)
```

This is an exploratory document intended to be used as 'scratch paper' for data analysis. Some of the code is left in for imitation as may be necessary. The comments throughout are meant to explain code choices and data summaries. Happy reading!

```{r, warning = FALSE}
### IMPORT DATA/CREATE SUBSETS  ###
###################################

data_wk2 <- readtext("/Users/ryanharris/Downloads/Bio2_Sp21_Week2_Survey_JM_RH.xlsx")
data_wk14 <- readtext("/Users/ryanharris/Downloads/Bio2_Sp21_Week14_Survey_JM_RH.xlsx")
data_wk14
## data_alv comes from a file created outside of R, included file (merged.xlsx) with rmd email
data_alv <- readtext("/Users/ryanharris/Desktop/merged.xlsx")

## each of the readtext() functions above are path specific to my machine
## to imitate, use data <- readtext("/path_to_files/data.xslx")

subset_wk2 <- data_wk2[ -c(1:3) ]
subset_wk14 <- data_wk14[ -c(1) ]

colnames(subset_wk2) <- c("Participant.ID", "q1.2", "q2.2", "q3.2", "lecture")
colnames(subset_wk14) <- c("Participant.ID", "q1.14", "q2.14", "q3.14", "lecture")

merged_data <- merge(subset_wk2, subset_wk14, by.x = "Participant.ID", by.y = "Participant.ID")
merged_data <- merged_data[ -c(9) ]

data_alv <- data_alv[-c(1)]
colnames(data_alv) <- c("Participant.ID", "survey", "q1", "q2", "q3", "lecture")
```

```{r, warning = FALSE}
## replace column values with numbers for easier data visuals

likert_num2 <- revalue(merged_data$q2.2, c("agree" = 5, "somewhat agree" = 4, "I don't know" = 3, "somewhat disagree" = 2, "disagree" = 1))
likert_num14 <- revalue(merged_data$q2.14, c("agree" = "5", "somewhat agree" = "4", "I don't know" = "3", "somewhat disagree" = "2", "disagree" = "1"))

merged_data <- merged_data[ -c(3, 7) ] ##these columns will be replaced with numbers
likert <- cbind(merged_data, likert_num2, likert_num14)

likert_cols <- likert[ -c(1,2,3,4,5,6) ]
likert_num <- data.frame(lapply(likert_cols,as.numeric))

likert_num$difference <- likert_num$likert_num14 - likert_num$likert_num2
likert <- cbind(likert, likert_num$difference)

colnames(likert) <- c("Participant.ID", "q1.2", "q3.2", "lecture.x", "q1.14", "q3.14", "likert_num2", "likert_num14", "difference")
col_order <- c("Participant.ID", "q1.2", "likert_num2", "q3.2", "lecture.x", "q1.14", "likert_num14", "q3.14", "difference")
likert <- likert[, col_order]
```

```{r, warning = FALSE}
## this is a more efficient way to create the subsets requested

increasing <- likert[ !(likert$difference <= 0 ), ]
#write.xlsx(increasing, file = "increasing")
decreasing <- likert[ !(likert$difference >= 0 ), ]
#write.xlsx(decreasing, file = "decreasing")
omit_2agree <- likert[ !(likert$difference == 0 & likert_num2 == 5 ), ]
#write.xlsx(omit_2agree, file = "no_both_agree")
```

```{r, warning = FALSE}
### FOR SUMMARY STATS ###
#########################

drops <- c("Participant.ID", "q1.2", "q3.2", "q1.14", "q3.14", "difference")
increasing_num <- increasing[ , !(names(increasing) %in% drops)]
decreasing_num <- decreasing[ , !(names(decreasing) %in% drops)]
omit2_num <- omit_2agree[ , !(names(omit_2agree) %in% drops)]

increasing_num <- data.frame(lapply(increasing,as.numeric))
decreasing_num <- data.frame(lapply(decreasing,as.numeric))
omit_2agree_num <- data.frame(lapply(omit_2agree,as.numeric))
```

```{r, include = FALSE}
### SUMMARY STATS AND FREQUENCY DIAGRAMS  ###
#############################################

### TOTAL SUMMARY ###
summary(likert_num)
ggplot(likert_num) +
  geom_bar(aes(likert_num2))
ggplot(likert_num) +
  geom_bar(aes(likert_num14))
ggplot(likert_num) +
  geom_bar(aes(difference)) +
  scale_x_continuous(breaks = pretty_breaks())

col_n <- c(3,7,9)

### OMIT 2 AGREE SUMMARY  ###
omit_sum <- omit_2agree_num[,col_n]
summary(omit_sum)
ggplot(omit_sum) +
  geom_bar(aes(likert_num2))
ggplot(omit_sum) +
  geom_bar(aes(likert_num14))
ggplot(omit_sum) +
  geom_bar(aes(difference)) +
  scale_x_continuous(breaks = pretty_breaks())

### INCREASING SUMMARY  ###
increasing_sum <- increasing_num[,col_n]
summary(increasing_sum)
ggplot(omit_sum) +
  geom_bar(aes(likert_num2))
ggplot(omit_sum) +
  geom_bar(aes(likert_num14))
ggplot(omit_sum) +
  geom_bar(aes(difference)) +
  scale_x_continuous(breaks = pretty_breaks())

### DECREASING SUMMARY  ###
decreasing_sum <- decreasing_num[,col_n]
summary(decreasing_sum)
ggplot(decreasing_sum) +
  geom_bar(aes(likert_num2))
ggplot(decreasing_sum) +
  geom_bar(aes(likert_num14))
```

```{r}
alluvial1 <- omit_2agree[ , !(names(omit_2agree) %in% c("Participant.ID", "q1.2", "q3.2", "q1.14", "q3.14", "difference"))]
```

```{r}
### REVERT LIKERT COLUMNS TO RESPONSES  ###
###########################################

dp <- c("likert_num2", "lecture.x", "likert_num14")
likert_org2 <- revalue(alluvial1$likert_num2, c("5" = "agree", "4" = "somewhat agree", "3" = "I don't know", "2" = "somewhat disagree", "1" = "disagree"))
likert_org14 <- revalue(alluvial1$likert_num14, c("5" = "agree", "4" = "somewhat agree", "3" = "I don't know", "2" = "somewhat disagree", "1" = "disagree"))
alluvial2 <- cbind(likert_org2, likert_org14)

likert_inc2 <- revalue(increasing$likert_num2, c("4" = "somewhat agree", "3" = "I don't know", "2" = "somewhat disagree", "1" = "disagree"))
likert_inc14 <- revalue(increasing$likert_num14, c("5" = "agree", "4" = "somewhat agree", "3" = "I don't know", "2" = "somewhat disagree"))
increasing2 <- cbind(increasing, likert_inc2, likert_inc14)
increasing2 <- increasing2[ , !(names(increasing2) %in% c("Participant.ID", "q1.2", "likert_num2", "q3.2", "lecture.x", "q1.14", "likert_num14", "q3.14", "difference")) ]
```

```{r}
### RANDOMIZE INCREASING SUBSET ###
###################################

# Data is randomized by row. I created two new dataframes - rand_increasing and rand_inc_censor. The first is for reference (so that we may un-randomize the dataframe or identify participants as needed). The second is the censored version (for Sheelta). 

# both dataframes written to spreadsheets and attached to email with this Rmd file.

rand <- sample(nrow(increasing))
rand_increasing <- increasing[rand,]
rand_inc_censor <- rand_increasing[ , !(names(rand_increasing) %in% c("Participant.ID", "lecture.x"))]
#write.xlsx(rand_increasing, file = "increasing_uncensored.xlsx")
#write.xlsx(rand_inc_censor, file = "increasing_censored.xlsx")
```

The first alluvial diagram includes everything asked of me excepting the order of the strata.  The second diagram includes everything asked of me excepting the labelling of the strata with words instead of numbers.  I am having trouble implementing both simultaneously, but I will continue to look at it.  The next two alluvial diagrams are the same, but with the increasing subset.

```{r, warning = FALSE}
### MERGED DATA WITHOUT BOTH AGREE  ###
#######################################

ggplot(alluvial2, aes(axis1 = likert_org2, axis2 = likert_org14)) +
  geom_alluvium(aes(fill = likert_org2), reverse = FALSE) +
  geom_stratum(reverse = FALSE) +
  scale_fill_manual(values=c("#FF0000", "#eb8686", "#808080", "#86a6d9", "#6271e3")) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE)

ggplot(omit_2agree, aes(axis1 = likert_num2, axis2 = likert_num14)) +
  geom_alluvium(aes(fill = likert_num2), reverse = FALSE) +
  geom_stratum(reverse = FALSE) +
  scale_fill_manual(values=c("#FF0000", "#eb8686", "#808080", "#86a6d9", "#6271e3")) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE)
```

```{r, warning = FALSE}
### INCREASING SUBSET ###
#########################

ggplot(increasing2, aes(axis1 = likert_inc2, axis2 = likert_inc14)) +
  geom_alluvium(aes(fill = likert_inc2), reverse = FALSE) +
  geom_stratum(reverse = FALSE) +
  scale_fill_manual(values=c("#FF0000", "#eb8686", "#808080", "#86a6d9", "#6271e3")) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE)

ggplot(increasing, aes(axis1 = likert_num2, axis2 = likert_num14)) +
  geom_alluvium(aes(fill = likert_num2), reverse = FALSE) +
  geom_stratum(reverse = FALSE) +
  scale_fill_manual(values=c("#FF0000", "#eb8686", "#808080", "#86a6d9", "#6271e3")) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE)
```
Any diagrams found below here were not included in the top two choices for the 2-panel figure for the grant report.  I can implement the same requests from the above subsets with any of the data as needed.

```{r, warning = FALSE}
### DECREASING SUBSET ###
#########################

## color by week 2 responses
ggplot(decreasing, aes(axis1 = likert_num2, axis2 = likert_num14)) +
  geom_alluvium(aes(fill = likert_num2), reverse = FALSE) +
  geom_stratum(reverse = FALSE) +
  scale_fill_manual(values=c("#FF0000", "#eb8686", "#808080", "#86a6d9", "#6271e3")) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE)
```

```{r, warning = FALSE}
### MERGED/ALV DATA ###
#######################

ggplot(merged_data, aes(axis1 = likert_num2, axis2 = likert_num14)) +
  geom_alluvium(aes(fill = likert_num2), reverse = FALSE) +
  geom_stratum() +
  stat_stratum(reverse = FALSE) +
  scale_fill_manual(values=c("#FF0000", "#eb8686", "#808080", "#86a6d9", "#6271e3")) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE)
```

```{r}
alluvial1 <- omit_2agree[ , !(names(omit_2agree) %in% c("Participant.ID", "q1.2", "q3.2", "q1.14", "q3.14", "difference"))]
omit_2agree
```